---
title: "p8105_hw3_msk2265"
author: "Mirah"
date: "`r Sys.Date()`"
output: github_document
---

## Problem 1 
```{r setup, include=FALSE}

library(tidyverse)
library(ggridges)
library(p8105.datasets)
data("brfss_smart2010")
```

**First, do some data cleaning:**

**format the data to use appropriate variable names;**
  - **focus on the “Overall Health” topic**
  - **include only responses from “Excellent” to “Poor”**
  - **organize responses as a factor taking levels ordered from “Poor” to “Excellent”**
  
```{r}
#cleaning data

order_vector = c("Poor", "Excellent") #making order for factor variable

brfss_df = brfss_smart2010 %>%
  janitor::clean_names() %>% 
  filter(topic== "Overall Health", response == "Poor" | response == "Excellent") %>% 
  mutate(response = factor(response, levels = order_vector))
  #I can remove several columns to make the data cleaner, but that is not in the instructions so I am not sure if I should be doing that
```

**Using this dataset, do or answer the following (commenting on the results of each):**

In 2002, which states were observed at 7 or more locations? What about in 2010?

```{r}
#creating a data frame to calculate how many state were observed at 7 or more locations
brfss_df_2002 = brfss_df %>% 
  filter(year == 2002) %>% 
  group_by(locationabbr) %>% #group by state
  summarize(unique_location = n_distinct(locationdesc)) %>% #unique locations in the state
  filter(unique_location > 7) #filter for more than 7 unique locations
count(brfss_df_2002) # counting

#same thing for 2010
brfss_df_2010 = brfss_df %>% 
  filter(year == 2010) %>% 
  group_by(locationabbr) %>% #group by state
  summarize(unique_location = n_distinct(locationdesc)) %>% #unique locations in the state
  filter(unique_location > 7) #filter for more than 7 unique locations
count(brfss_df_2010) # counting

```
In 2002,  there were 3 states that were observed in 7 or more locations, and in 2010 there were 11 states that were observed in 7 or more unique locations. 

**Construct a dataset that is limited to Excellent responses, and contains, year, state, and a variable that averages the data_value across locations within a state. Make a “spaghetti” plot of this average value over time within a state (that is, make a plot showing a line for each state across years – the geom_line geometry and group aesthetic will help).**
```{r}
#creating the dataset
brfss_state_averages = brfss_df %>% 
  filter(response == "Excellent") %>% 
  group_by(locationabbr, year) %>% #need to group to calculate mean
  mutate(data_value_average = mean(data_value)) %>% 
  ungroup(locationabbr, year) %>% #ungroup
  select(year, locationabbr, data_value_average) %>% 
  distinct() #getting rid of repeats since each state is listed multiple times
  
#create plot
ggplot(brfss_state_averages, aes(x=year, y= data_value_average, color = locationabbr)) + geom_line()
```

**Make a two-panel plot showing, for the years 2006, and 2010, distribution of data_value for responses (“Poor” to “Excellent”) among locations in NY State**
```{r}
brfss_ny = brfss_df %>% 
  filter(locationabbr == "NY", year == 2006 | year == 2010)

ggplot(brfss_ny, aes(x=data_value, fill = response)) + 
  geom_histogram() +
  facet_grid(. ~ year)
```

## Problem 3

***Load, tidy, merge, and otherwise organize the data sets. Your final dataset should include all originally observed variables; exclude participants less than 21 years of age, and those with missing demographic data; and encode data with reasonable variable classes (i.e. not numeric, and using factors with the ordering of tables and plots in mind).***

```{r}
#specifying how to level education
education_level = c("Less than high school", "High school equivalent", "More than high school")
#loading data
 nhanes_demographic= read_csv("./Data/nhanes_covar.csv", skip=4, na = "NA" ) %>% 
  janitor::clean_names() %>% 
  drop_na() %>% # droppong rows that are missing demographic data
  mutate(
    sex = case_match(
      sex, 
      1 ~ "male",
      2 ~ "female"
    )) %>% #coding sex to character
   mutate(
     education= factor(case_match(
       education,
       1 ~ "Less than high school",
       2 ~ "High school equivalent",
       3 ~ "More than high school"
     ), levels = education_level)
   ) %>% 
   filter(age>20)#age less than 21 filtered out so includes 21
 
  
nhanes_accel = read_csv("./Data/nhanes_accel.csv", na = "0") %>% #assuming 0 means that no data waas collected at that time point
  janitor::clean_names()

#merging datasets

nhanes_df = left_join(nhanes_demographic, nhanes_accel, by = "seqn")
```


```{r}
#covert to long data? not sure if this is needed but doing it anyways just in case
nhanes_long_df = (
  pivot_longer(nhanes_df,
               min1:min1440,
               names_to = "minutes",
               names_prefix = "min",
               values_to = "MIM")) %>% 
  mutate(
    minutes = as.numeric(minutes)
  )

```


***Produce a reader-friendly table for the number of men and women in each education category, and create a visualization of the age distributions for men and women in each education category. Comment on these items.***
```{r}
#table with number of men and women in each education category
nhanes_df %>% 
  group_by(education, sex) %>% 
  summarize (n_obs = n())

ggplot(nhanes_df, aes(x=age, fill = sex))+
  geom_histogram() +
  facet_grid(. ~ education)
```


***Traditional analyses of accelerometer data focus on the total activity over the day. Using your tidied dataset, aggregate across minutes to create a total activity variable for each participant. Plot these total activities (y-axis) against age (x-axis); your plot should compare men to women and have separate panels for each education level. Include a trend line or a smooth to illustrate differences. Comment on your plot.***
```{r}
#this is easier to do with the wide dataset

nhanes_df = nhanes_df %>% 
  mutate(activity = rowMeans(select(.,min1:min1440), na.rm=TRUE)) #adding activity row by calculating the average time for each row, ignoring the 0 values that were not collected.

ggplot(nhanes_df, aes(x = age, y = activity, color = sex)) +
  geom_point(alpha = .5)+
  geom_smooth(se= FALSE)+
  facet_grid(.~ education)

```


Accelerometer data allows the inspection activity over the course of the day. Make a three-panel plot that shows the 24-hour activity time courses for each education level and use color to indicate sex. Describe in words any patterns or conclusions you can make based on this graph; including smooth trends may help identify differences.